Probability of causal contacts between interstellar civilizations through Monte Carlo simulations

### 2020-05-26 submit_v03/manuscript.tex
segunda lectura

\begin{figure}[!t]
   %
   \centering
   %
   \includegraphics[width=0.5\textwidth]{F_scheme.pdf}
   %
   \caption{
   %
Space--time diagrams showing a schematic representation of the different stages
in the development of a constrained causal contact node (\ceti{}).
%
Panel (a) represents the region $R$ in space and time which is causally
connected to the emitter (E, left vertex), following an ``A'' type
event (``Awakening'') in which the node acquires the communication
capability.
%
The sphere of first contact (SFC) is the sphere centered on the emitter that
grows until its radius reaches the D$_{\rm max}$ distance, in which the
power of the signal would equal the detectability threshold.
%
In the Figure this sphere is represented by the left triangle of the region
$R$.
%
The surface of last contact (SLC) is another sphere that grows from a ``D''
event (``Doomsday'').
%
The region which is causally connected to the emitter is then limited by these
two spheres, and has the shape of a sphere or of a spherical shell,
depending on the time.
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b).
%
The receiver node E$_1$ can listen signals from the emitter node E$_2$, from
the ``Contact'' event (t=C$_{12}$) up to the ``Blackout'' event
(t=B$_{12}$).
%
Similarly, the receiver \ceti{} E$_2$ can listen signals from the emitter
\ceti{} E$_1$, from t=C$_{21}$ up to t=B$_{21}$. 
%
} \label{F_scheme}
%
\end{figure}

@@

\ffn{1} Space—time diagrams, where time and space are represented on
the horizontal and vertical axes, respectively, are suitable tools to
describe the causal connections among different nodes.
%
We illustrate in the Fig.~\ref{F_scheme} the schematic representation
of the region causally connected to a given \ceti{}.
%
For the sake of simplicity, we show in the plot the light travel
distance, i.e., the distance traversed by any signal spreading from
the emitter at light speed (for example, any electromagnetic signal).
%
In this scheme, a light pulse would follow a trajectory represented by
a line at 45 degrees from the axes, given the units of time and
space axes are years and light years, respectively.
%
Panel (a) of the Figure represents the region $R$ (shaded polygon)
which is causally connected to the emitter (left vertex).
%
This region develops after an event (hereafter dubbed A-type event) in
which the node acquires the communication capability, or becomes
''active''.
%
The sphere of first contact (SFC) is centred on the emitter,
represented in the Figure by the left angle of the region $R$.
%
This sphere grows until its radius reaches the D$_{\rm max}$ distance, in
which it would be no longer detectable due to the decrease in the
energy per unit area, which falls under the assumed detectability
threshold.
%
Similarly, the surface of last contact (SLC) is another sphere that
grows from an event in which the nodes ceases to possess the
communication capability (D type event), and carries the last signal
produced by the emitter.
%
The region which is causally connected to the emitter is then limited
by these two spheres and therefore has the shape of a filled sphere
(before the D event) or of a spherical shell (after the D event).
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b) of the Figure.
%
In this scheme, the entire lifetime of a \ceti{} can be represented by
a polygon, as shown in the Figure.
%
The polygon is limited in the spatial axis by the maximum distance of
the signal $D_{\rm max}$ and in the time axis by the time span between the
A-type event (concave vertex) and the D-type event (convex vertex).
%
This representation allows to visualize the important events that
result from the presence and communications of two \cetis{}.
%
The points in time where the \cetis{} acquire their communicating
capacity are dubbed ‘’Awakening’’ events ($A_1$ and $A_2$ in the
Figure).
%
Similarly, the instants in time where the \cetis{} lose their
communicating capacity are dubbed ‘’Doomsday’’ events ($D_1$ and $D_2$
in the Figure).
%
The points in space--time where the first contact is produced for each
one of the \cetis{}, are defined as ''Contact'' events, shown as $C_1$
and $C_2$.
%
Finally, the two points where the contact is lost for each one the
\cetis{}, are denominated ''Blackout'' events, presented as $B_1$ and
$B_2$ in the Figure.
%
The receiver node E$_1$ can listen signals from the emitter node
E$_2$, from the first contact event (t=C$_{12}$) until the last
one (t=B$_{12}$).
%
Similarly, the receiver node E$_2$ can listen signals from the emitter
node E$_1$, from t=C$_{21}$ up to t=B$_{21}$.
%
The time intervals for the open communication channel are determined
by the ''C'' and ''B'' type events.
%
It is noteworthy the fact that there is a time delay between the
contact events of the two \cetis{} involved in this analysis, and also
a time delay between the two blackout events.
%
Therefore, the time range when a bidirectional contact is possible
occurs between the maximum time of the contact points and the minimum
time between the blackout points.

@@

In our experimental configuration the simulation starts assuming the stochastic process is already stable, and finishes before any galactic evolution effect could modify the fixed values of the variables.
%
Likewise, we assume that the probability for the appearance of a \ceti{} is homogeneous over the GHZ.
%
Although the Galaxy has a well-known spiral structure, the assumed sparsity of nodes (arguably, intelligent civilisations) does make that assumption tenable.
%
If the distribution of nodes is not sparse, as it could be the case if the spiral arms host most of the \cetis{}, then contacts would be more frequent between closely located nodes.
%
In such a case our results would underestimate the number of contacts between close pairs of nodes within the same spiral arm, and conversely, overestimate the number of contacts between separated nodes.
%
We also limit the possibilities of life or other types of civilisations to the usually stated hypothesis for the definition of
the GHZ \citep{dayal_habitability_2016, gonzalez_galactic_2001, lineweaver_galactic_2004, gonzalez_habitable_2005,
morrison_extending_2015, haqq-misra_evolution_2019, rahvar_cosmic_2016, gobat_evolution_2016, rahvar_cosmic_2016} and consider that habitability remains constant over time \citep[see, however, ][]{gonzalez_habitable_2005, dayal_habitability_2016, gobat_evolution_2016}.
%
For example, \citet{vukotic_astrobiological_2012} propose
probabilistic cellular automata model to explain the astrobiological
history of the Milky Way. % <<<< FEO
%
This means that we set aside civilisations that could survive in severe conditions or unstable systems, which would prevent the appearance of life as we know it.

We need to raise the scope of our results by stressing the fact that we are considering causal contacts instead of actual contacts.
%
By a causal contact we mean a configuration, in space and time, in which a node belongs to the light cone of another node.
%
This means that a causal contact does not need to be tied to any actual message emission or detection.
%
In more realistic scenarios, there are several sources of ‘’signal loss’’ with respect to the ideal case.
%
Among them, we can mention temporal and signal power aspects and direction dependent communication capabilties.
%
The results must then be interpreted as the case of ideal nodes, with a perfect efficiency in the emision and reception of signals.
%
The probabilities of contacts presented here are then upper limits to the number of communications between civilization s in the Galaxy, given the implemented hipotheses in the model.
%
The use of light cones as causal contact regions is inspired by the fact that light—speed traveling messengers like 
electromagnetic radiation, gravitational radiation or neutrinos are often considered as possible message carriers \citep{hippke_interstellar_2017, wright_how_2018}.
%
This excludes messages sent with mechanical means or physical objects \citep[e.g., ][]{Armstrong2013, barlow_galactic_2013}, or through some unknown technology that violates the known laws of physics.

As part of this benchmark, we assume that the capacity to emit and receive signals occur at the same time.
%
Although there are several reasons to think that this could not be the case, at large time scales it can be considered that both abilities occur roughly at coincident epochs.
%
In the ideal setup, this would be equivalent to nodes that send messages isotropically and scan the local skies on all directions with a perfect eficiency.
%
Another essential assumption is that all \cetis{} use the same signal power, so that there is a maximum distance out to which it can be detected.
%
% It is worth mentioning that we are considering in our experimental setup a system composed by several emitters and 
% receivers across the galaxy, under the hypothesis proposed before.
%
In such a system we compute probabilities of a \ceti{} making contact with another node, and the results must be interpreted as for a randomly chosen \ceti{}, given certain conditions of homogeneity on an ensemble of equivalent galaxies.
%
Then, we stress the fact that we are not centred on the probabilities for the Earth itself, but on a generalised \ceti{}. 
%
Regarding the extent of the signals, we know that the distance from which a signal from Earth could be detected using the current technology is of a few hundreds light years, given that the signal was sent to a specific direction.
%
It is straightforward to propose and implement a distribution of maximum distances, although this would increase the model complexity at the cost of a larger uncertainty or of another dimension on the hypothesis space.

@@

In our simulations, we assume the simplest configuration for the
growth of the sphere of first contact. For example, we discard the
possibility of stellar colonisation
\citep[e.g.][]{newman_galactic_1981, walters_interstellar_1980,
starling_virulence_2013, barlow_galactic_2012, jeong_large_2000,
maccone_mathematical_2011}. We also assume that communication is
equally likely in all directions, i.e., we assume isotropic
communication in all cases. Different communication efficiencies or
detection methods are straightforward to carry out in the simulations
for more detailed and complex scenarios. This approach, however, is
beyond this work because it obscures the experimental setup and make
the results less clear. The assumptions we accept imply that the
results are independent of whether intelligent agents are organic or
artificial. Moreover, the causes of the limited lifetime of a
civilization can be natural (astrophysical phenomena), caused by auto
destruction or by external factors, to name a few. However, we assume
that these events are sufficiently numerous in the Galaxy so that a
statistical model is plausible. The failure of this hypothesis would
imply extreme values for the parameters that represent the density of
nodes.

@@

The temporal structure of the process is defined by two distribution parameters. One of them represent the mean time
interval that a \ceti{} can emit and receive signals, and the other the mean time interval between the emergence of
consecutive \cetis{}. The spatial structure of the simulation is given by the size and shape of the Galactic Habitable
Zone and the maximum distance a signal can travel to be detected
(D$_{\rm max}$). The parameters for the temporal
distributions also determine the spatial properties, since the density of active \cetis{} in the galaxy depend on these
two parameters. For example, a small $\tau_a$ and a large $\tau_s$ will produce a galaxy densely populated with
\cetis{}. Also, some hypotheses regarding the shapes of the distributions of the temporal parameters must be made in
order to complete the simulation. \citet{forgan_spatiotemporal_2011} argues that the times at which different
civilisations become intelligent follow a Gaussian distribution, and then the distribution of inter-arrival times is an
inverse exponential. We assume that the distribution of the times of A events is a stationary Poisson process, and then
the distribution of the times between the appearance of new consecutive \cetis{} is exponential. Regarding the duration
of a \ceti{}, we propose that the distribution of the duration of \cetis{} is a stationary exponential distribution.
There are no clear arguments to conclude a statistical law for the later distribution, so that we propose it as a
working hipothesis.  A theoretical statistical distribution of the lifespan of civilisations would rely preferentially
on the basis of the underlying astrophysical and biological processes \citep{balbi_impact_2018}. For example,
\citet{maccone_lognormals_2014} argues that this distribution should be a log normal.

The power law and exponential statistical distributions are among the most common patterns found in natural phenomena.
%
For example, the distribution of the frequency of words in many languages is known to follow a Zipf’s law (which is a 
power law).
%
These distributions arise any time a phenomenon is characterised by commonly occurring small events and rarely produced
large events \citep[e.g. ][]{adamic_zipf_2000}.
%
Zipf law also describes population ranks of cities in various countries, corporation sizes, income rankings, ranks of
the number of people watching the same TV channel, etc.
%
The magnitudes of earthquakes, hurricanes, volcanic eruptions and floods; the sizes of meteorites or the losses caused
by business interruptions from accidents, are also well described by power laws
\citep{sornette_critical_2006}.
%
%% , benguigui_classificacion_2016}.
The power law behaviour has been observed in a variety of systems.
%
These include, for example, stock market fluctuations, sizes of computer files or word frequency in languages
\citep{mitzenmacher_brief_2004, newman_power_2005, simkin_theory_2006}. 
%
Power laws have also been widely used in
biological sciences.
%
Some examples are the analysis of connectivity patterns in metabolic networks \citep{jeong_large_2000} or the number of
species observed per unit area in ecology \citep{martin_origin_2006, frank_common_2009}.
%
More examples can be found in the literature 
\citep{martin_origin_2006, maccone_KLT_2010}
\citep{barabasi_scale_2009, maccone_evolution_2014}
\citep{maccone_lognormals_2014}

In this work, the exponential distribution of lifespan and waiting times is justified by considering the hypothesis that the process of appearance of life in the galaxy is homogeneous and stationary. That is, there is not a preferred region within the GHZ for the spontaneous appearance of life, and the emergence of a \ceti{} is independent of the existence of previous nodes in the galaxy. This is equivalent to proposing a Poisson process for the emergence of \cetis{}, given the relation between the number of events in time or space and the waiting time or separation, respectively
\citep[e.g., ][]{ross_simulation_2012}. That is, these are two alternative approaches to describing the same process, a Poisson distribution for the number of events implies an exponential distribution for their separations, and vice versa. It should be emphasised that the exponential laws used in this work are assumed as part of the working hypothesis, and instead of analysing results from a particular parameter chosen ad hoc, we explore the hypothesis space and analyse the impact of the values of these parameters on the results.

@@

In this Section we discuss the degree of complexity in the model, considering a compromise between the accuracy of the
model and the number of parameters that are free or with a high uncertainty.
%
Firstly, we emphasise that the odds of a causal contact between two \cetis{} should not be considered as the odds of a
contact between two intelligent civilisations, and in fact the latter could be much lesser than the former.
%
Indeed, in order to establish a contact between any two entities, a minimum degree of compatibility must be accomplished
without any previous agreement, making the possibility of a contact with a message that could be deciphered highly rare
\citep[see e.g. ][]{forgan_collimated_2014}.
%
Besides the trade--off between the simplicity and the complexity of the experiments, further analysis could be performed
following this framework in order to explore possible implications of the results for more detailed configurations.
%
For example, the communication method (isotropic, collimated, serendipitous) can affect the observables, making it
necessary to implement a correction factor.
%
Taking this into account, our results regarding the probability of causal contact should be considered as upper limits
for effective contacts, since they depend on the efficiency of both the emitter and the receiver to broadcast and scan
the sky for intelligent signals, respectively \citep{grimaldi_signal_2017}.
%
A correction by coverage ratio in the detection and by a targeting ratio in the emission could be easily implemented in
the simulation, although the effect of reducing the probability of contact is basically the product of the efficiency
ratios and thus such implementation is not necessary.
%
Therefore, the values of the probabilities could be modified by a constant correction factor equal to the combined
emission/reception efficiency \citep{smith_broadcasting_2009, anchordoqui_upper_2019, forgan_collimated_2014} or for
beam-like transmissions \citep{grimaldi_signal_2017}. 
%
Other considerations include the effects of alignments, the use of stars as sources or amplifiers \citep{Edmondson2003,
borra_searching_2012}, the nature of the message carrier.
%
Another improvement could be the use of a spatial distribution that resembles the spiral shape and the width of the disc
of the Galaxy.
%
Regarding the distributions of the parameters, different distribution functions for the mean lifetime of \cetis{} can be
implemented as alternatives.
%
Finally, regarding the degree of idealisation, the model admits different efficiencies for the sphere of the causal
region of each \ceti{}, featuring different searching strategies \citep{hippke_interstellar_2017}. 
%
It is also possible to consider that $D_{\rm max}$ is different for different \cetis{}.
%
For example, a power law where powerful emissions are rare and low emissions are common could be an improvement to the
model.
%
In this work we chose not to implement this for the sake of simplicity.
%
Finally, the role of the message contents could influence on the lifespan of a \ceti{} that receives a message, although
the implementation of such behaviour would increase the number of free parameters and would be more speculative.

@@

A discrete event simulation is performed for a given model $M(\tau_a,
\tau_s, D_{\rm max})$ by keeping track of a set of variables that change each time an event happens. 
%
In addition, the model comprises elements that are fixed for all simulations, for example, the functional forms of the statistical distributions and the adopted values of particular variables (see Table XXX).
%
The main variables that follow the evolution of the simulation are: the positions of stars, which are sampled randomly within the GHZ; the time of the awakening of each \ceti{} (A event); and the time of the vanishing of each \ceti{} (D event). 
%
The variables that can be deduced from the previous ones include: the number of \cetis{} in casual contact with at least another \ceti{} at a given time; the number of \cetis{} as a function of time; the number of \cetis{} that receive a message at least one time; the number of \cetis{} that receive a message at least one time and successfully deliver an answer; and the number distribution of waiting time to receive a message.
%
All these quantities are updated each time one of the four events (A, B, C, D) occurs.

@@

Since the method is based on events, the first step is to define the relationships between them and how events trigger changes on the state of the system.
%
Once an event is produced, the system is updated by saving the corresponding change in the related variables.
%
For example, a B-type event diminishes the number of active \cetis{} by one, and also could affect the number of ongoing communications.
%
We consider a first approach to the galactic habitable zone as a 2-dimensional annular region.
%
The adopted values for the geometry of the GHZ are 2 kpc and 8 kpc, for the inner and outer radius, respectively \citep{lineweaver_galactic_2004}. 
% ojo, ese paper dice 7-9 kpc, o sea tiene un limite superior mucho
% mas chico...
%
This simple model does not take into account the variations in stellar density given by the spiral structure of the Galaxy.
%
The main task in setting a DE simulation is to define the events and what changes do these events produce on the system.
%
Although there are several possible approaches, we chose to follow the evolution of the system according to the following events:

@@

We have chosen this nomenclature so that it can be easily remembered
from the initials A, B, C, D.
%
The \blackout is produced when one of the two \cetis{} halts in its
capability of emitting and receiving signals.
%
It can occur at the same time the \doomsday or before.
%
Similarly, a \ccontact can be produced at the time of the \aawakening
or later, and there can be several \contacts for the same node.
%
The time marks for each event (A and D for each \ceti{} and C and B
for each causal contact) are stored as a result of the simulation.
%
Also, the list of active \cetis{} is obtained as a function of time.
%
%Some fixed variables must be set in order to carry out the simulation:
%the size of the Galactic Habitable Zone, the mean lifetime of a
%\ceti{}, $\tau_s$, the mean waiting time for the appearance of another
%\ceti{}, $\tau_a$, and the maximum distance a message can be detected.

>>>>

\begin{figure}[!t]
   %
   \centering
   %
   \includegraphics[width=0.5\textwidth]{F_scheme.pdf}
   %
   \caption{
   %
Space--time diagrams showing a schematic representation of the different stages
in the development of a constrained causal contact node (\ceti{}).
%
Panel (a) represents the region $R$ in space and time which is causally
connected to the emitter (E, left vertex), following an ``A'' type
event (``Awakening'') in which the node acquires the communication
capability.
%
The sphere of first contact (SFC) is the sphere centered on the emitter that
grows until its radius reaches the D$_{\rm max}$ distance, in which the
power of the signal would equal the detectability threshold.
%
In the Figure this sphere is represented by the left triangle of the region
$R$.
%
The surface of last contact (SLC) is another sphere that grows from a ``D''
event (``Doomsday'').
%
The region which is causally connected to the emitter is then limited by these
two spheres, and has the shape of a sphere or of a spherical shell,
depending on the time.
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b).
%
The receiver node E$_1$ can listen signals from the emitter node E$_2$, from
the ``Contact'' event (t=C$_{12}$) up to the ``Blackout'' event
(t=B$_{12}$).
%
Similarly, the receiver node E$_2$ can listen signals from the emitter
node E$_1$, from t=C$_{21}$ up to t=B$_{21}$. 
%
} \label{F_scheme}
%
\end{figure}

@@

\ffn{1} Space—time diagrams, where time and space are represented on
the horizontal and vertical axes, respectively, are suitable tools to
describe the causal connections among different nodes.
%
We illustrate in the Fig.~\ref{F_scheme} the schematic representation
of the region causally connected to a given \ceti{}.
%
For the sake of simplicity, we show in the plot the light travel
distance, i.e., the distance traversed by any signal spreading from
the emitter at light speed (for example, any electromagnetic signal).
%
In this scheme, a light pulse would follow a trajectory represented by
a line at 45 degrees from the axes, given the units of time and
space axes are years and light years, respectively.
%
Panel (a) of the Figure represents the region $R$ (shaded polygon)
which is causally connected to the emitter (left vertex).
%
This region develops after an event (hereafter dubbed A-type event) in
which the node acquires the communication capability, or becomes
''active''.
%
The sphere of first contact (SFC) is centred on the emitter,
represented in the Figure by the left angle of the region $R$.
%
This sphere grows until its radius reaches the D$_{\rm max}$ distance, in
which it would be no longer detectable due to the decrease in the
energy per unit area, which falls under the assumed detectability
threshold.
%
Similarly, the surface of last contact (SLC) is another sphere that
grows from an event in which the nodes ceases to possess the
communication capability (D type event), and carries the last signal
produced by the emitter.
%
The region which is causally connected to the emitter is then limited
by these two spheres and therefore has the shape of a filled sphere
(before the D event) or of a spherical shell (after the D event).
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b) of the Figure.
%
In this scheme, the entire lifetime of a \ceti{} can be represented by
a polygon, as shown in the Figure.
%
The polygon is limited in the spatial axis by the maximum distance of
the signal $D_{\rm max}$ and in the time axis by the time span between the
A-type event (concave vertex) and the D-type event (convex vertex).
%
This representation allows to visualize the important events that
result from the presence and communications of two \cetis{}.
%
The points in time where the \cetis{} acquire their communicating
capacity are dubbed ``Awakening'' events ($A_1$ and $A_2$ in the
Figure).
%
Similarly, the instants in time where the \cetis{} lose their
communicating capacity are dubbed ‘’Doomsday’’ events ($D_1$ and $D_2$
in the Figure).
%
The points in space--time where the first contact is produced for each
one of the \cetis{}, are defined as ``Contact'' events, shown as $C_1$
and $C_2$.
%
Finally, the two points where the contact is lost for each one the
\cetis{}, are denominated ``Blackout'' events, presented as $B_1$ and
$B_2$ in the Figure.
%
The receiver node E$_1$ can listen signals from the emitter node
E$_2$, from the first contact event (t=C$_{12}$) until the last
one (t=B$_{12}$).
%
Similarly, the receiver node E$_2$ can listen signals from the emitter
node E$_1$, from t=C$_{21}$ up to t=B$_{21}$.
%
The time intervals for the open communication channel are determined
by the ``C'' and ``B'' type events.
%
It is noteworthy the fact that there is a time delay between the
contact events of the two \cetis{} involved in this analysis, and also
a time delay between the two blackout events.
%
Therefore, the time range when a bidirectional contact is possible
occurs between the maximum time of the contact points and the minimum
time between the blackout points.

@@

In our experimental configuration the simulation starts assuming that the
stochastic process is already stable, and finishes before any galactic
evolution effect could modify the fixed values of the variables.
%
Likewise, we assume that the probability for the appearance of a \ceti{} is
homogeneous over the GHZ.
%
Although the Galaxy has a well-known spiral structure, the nodes are assumed
to be sparse (otherwise the Galaxy would be full of life) and the spiral
structure would not, in that case, produce significant differences.
%
If the distribution of nodes is not sparse, as it could be the case if the
spiral arms host most of the \cetis{}, then contacts would be more frequent
between closely located nodes.
%
In such a case our results underestimate the number of contacts between
close pairs of nodes within the same spiral arm, and conversely, overestimate
the number of contacts between separated nodes.
%
We also limit the possibilities of life or other types of civilisations to the
usually stated hypotheses for the definition of the GHZ
\citep{dayal_habitability_2016, gonzalez_galactic_2001,
lineweaver_galactic_2004, gonzalez_habitable_2005, morrison_extending_2015,
haqq-misra_evolution_2019, rahvar_cosmic_2016, gobat_evolution_2016,
rahvar_cosmic_2016} and consider that habitability remains constant over time
\citep[see, however, ][]{gonzalez_habitable_2005, dayal_habitability_2016,
gobat_evolution_2016}.
%
This means that we set aside civilisations that could survive in severe
conditions or unstable systems, which would prevent the appearance of life as
we know it.

We stress the fact that we are
considering causal contacts instead of actual contacts.
%
In more realistic scenarios, there are several sources of ``signal loss'' with
respect to the ideal case.
%
Among them, we can mention temporal and signal power aspects and direction
dependent communication capabilities.
%
The results must then be interpreted as the case of ideal nodes, with a perfect
efficiency in the emision and reception of signals.
%
The probabilities of contacts presented here are then upper limits to the
number of communications between civilizations in the Galaxy, given the
implemented hipotheses in the model.
%
The use of light cones as causal contact regions is inspired by the fact that
light-speed traveling messengers like electromagnetic radiation, gravitational
radiation or neutrinos are often considered as possible message carriers
\citep{hippke_interstellar_2017, wright_how_2018}.
%
This excludes messages sent with mechanical means or physical objects
\citep[e.g., ][]{Armstrong2013, barlow_galactic_2013}, or through some unknown
technology that violates the known laws of physics.

As part of this benchmark, we assume that the capacity to emit and receive
signals occur at the same time.
%
Although there are several reasons to think that this could not be the case, at
large time scales it can be considered that both abilities occur roughly at
coincident epochs.
%
In the ideal setup, this would be equivalent to nodes that send messages
isotropically and scan the local skies on all directions with a perfect
eficiency.
%
Another essential assumption is that all \cetis{} use the same signal power, so
that there is a maximum distance out to which it can be detected.
%
% It is worth mentioning that we are considering in our experimental setup a
% system composed by several emitters and  receivers across the galaxy, under
% the hypothesis proposed before.
%
In such a system we compute probabilities of a random node making contact with
another node, i.e., they are not specific for the case of contacts with the
Earth.
%
Regarding the extent of the signals, we know that the distance from which a
signal from Earth could be detected using the current technology is of a few
parsecs, given that the signal was sent to a specific direction.
%
It is straightforward to propose and implement a distribution of maximum
distances, although this would increase the model complexity at the cost of a
larger uncertainty.

@@

In our simulations, we assume the simplest configuration for the
growth of the sphere of first contact.
%
In particular, we do not consider the
possibility of stellar colonisation
\citep[e.g.][]{newman_galactic_1981, walters_interstellar_1980,
starling_virulence_2013, barlow_galactic_2012, jeong_large_2000,
maccone_mathematical_2011}.
%
We also assume that communication is
equally likely in all directions, i.e., we assume isotropic
communication in all cases.
%
Different communication efficiencies or
detection methods are straightforward to carry out in the simulations
for more detailed and complex scenarios.
%
This approach, however, is
beyond this work because it obscures the experimental setup and make
the results less clear.
%
The assumptions we accept imply that the
results are independent of whether intelligent agents are organic or
artificial. Moreover, the causes of the limited lifetime of a
civilization can be natural (astrophysical phenomena), caused by auto
destruction or by external factors, to name a few. However, we assume
that these events are sufficiently numerous in the Galaxy so that a
statistical model is plausible. The failure of this hypothesis would
imply extreme values for the parameters that represent the density of
the nodes (i.e., $\left< \tau_a \right>$).

@@

The temporal structure of the process is defined by two distribution
parameters. One of them represent the mean time interval that a node can
emit and receive signals (its lifetime), and the other the mean time interval between the
emergence of consecutive \cetis{}.
%
The spatial structure of the simulation is
given by the size and shape of the Galactic Habitable Zone and the maximum
distance a signal can travel to be detected (D$_{\rm max}$). The parameters for
the temporal distributions also determine the spatial properties, since the
density of active \cetis{} in the galaxy depend on these two parameters. For
example, small $\tau_a$ and a large $\tau_s$ will produce a densely
populated galaxy (in this context, a galaxy is an element in a 
statistical ensemble, not the Milky Way).
%
Also, some hypotheses regarding the shapes of the
distributions of the temporal parameters must be made in order to complete the
simulation. \citet{forgan_spatiotemporal_2011} argues that the times at which
different civilisations become intelligent follow a Gaussian distribution, and
then the distribution of inter-arrival times is an inverse exponential. We
assume that the distribution of the times of A events is a stationary Poisson
process, and then the distribution of the times between the appearance of new
consecutive nodes is exponential. Regarding the duration of a node, we
propose that its distribution is a stationary
exponential distribution.  There are no clear arguments to conclude a
statistical law for the later distribution, so that we propose it as a working
hipothesis.  
%
This heuristic does not make any consideration about the origin of life, although
different approaches are possible.
%
For example,
\citet{maccone_lognormals_2014} argues that this distribution should be a log-normal. 
%
Preferentially, a theoretical statistical distribution of the lifespan of
civilisations would rely on the basis of the underlying
astrophysical and biological processes \citep{balbi_impact_2018}.

The power law and exponential statistical distributions are among the most
common patterns found in natural phenomena.
%
For example, the distribution of the frequency of words in many languages is
known to follow the law of Zipf (which is a power law).
%
These distributions arise any time a phenomenon is characterised by commonly
occurring small events and rarely produced large events \citep[e.g.
][]{adamic_zipf_2000}.
%
Zipf law also describes population ranks of cities in various countries,
corporation sizes, income rankings, ranks of the number of people watching the
same TV channel, etc.
%
The magnitudes of earthquakes, hurricanes, volcanic eruptions and floods; the
sizes of meteorites or the losses caused by business interruptions from
accidents, are also well described by power laws
\citep{sornette_critical_2006}.
%
%% , benguigui_classificacion_2016}.
%The power law behaviour has been observed in a variety of systems.
%
Additional examples include stock market fluctuations, sizes of computer files
or word frequency in languages \citep{mitzenmacher_brief_2004,
newman_power_2005, simkin_theory_2006}. 
%
Power laws have also been widely used in biological sciences.
%
Some examples are the analysis of connectivity patterns in metabolic networks
\citep{jeong_large_2000} or the number of species observed per unit area in
ecology \citep{martin_origin_2006, frank_common_2009}.
%
More examples can be found in the literature \citep{martin_origin_2006,
maccone_KLT_2010} \citep{barabasi_scale_2009, maccone_evolution_2014}
\citep{maccone_lognormals_2014}
%
This distribution family is suitable for the statistical description
of the $D_{\rm mar}$ parameter, although in this work we assume a uniform
distribution for simplicity.
%
The power law family of functions is also a good candidate for the description of
the temporal variables in the model.
%
However, we prefer the exponential model.
%
The exponential distribution of lifespan and waiting times is justified by
considering the hypothesis that the process of appearance of life in the galaxy
is homogeneous and stationary.
%
That is, there is not a preferred region within
the GHZ for the spontaneous appearance of life, and the emergence of a node is
independent of the existence of previous nodes in the galaxy.
%
These seem to be simple conditions, and allow to propose a distribution family
without knowing the details of the underlying process.
%
The exponential distribution for the separations in time is equivalent to
proposing a Poisson process for the emergence of nodes, given the relation
between the number of events in time or space and the waiting time or
separation, respectively \citep[e.g., ][]{ross_simulation_2012}.
%
That is, these
are two alternative approaches to describing the same process, a Poisson
distribution for the number of events implies an exponential distribution for
their separations, and vice versa.
%
It should be emphasised that the exponential
laws used in this work are assumed as part of the working hypothesis, and
instead of analysing results from a particular parameter chosen ad hoc, we
explore the hypothesis space and analyse the impact of the values of these
parameters on the results.

@@

In this Section we discuss the degree of complexity in the model, considering a
compromise between the accuracy of the model and the number of parameters that
are free or with a high uncertainty.
%
Firstly, we emphasise that the odds of a causal contact between two nodes
should not be considered as the odds of a contact between two intelligent
civilisations, and in fact the latter could be much lesser than the former.
%
Indeed, in order to establish a contact between any two entities, a minimum
degree of compatibility must be accomplished without any previous agreement,
making the possibility of a contact with a message that could be deciphered
highly rare \citep[see e.g. ][]{forgan_collimated_2014}.
%
Besides the trade--off between the simplicity and the complexity of the
experiments, further analysis could be performed following this framework in
order to explore possible implications of the results for more detailed
configurations.
%
For example, the communication method (isotropic, collimated, serendipitous)
can affect the observables, making it necessary to implement a correction
factor.
%
Taking this into account, our results regarding the probability of causal
contact should be considered as upper limits for effective contacts, since they
depend on the efficiency of both the emitter and the receiver to broadcast and
scan the sky for intelligent signals, respectively
\citep{grimaldi_signal_2017}.
%
A correction by coverage ratio in the detection and by a targeting ratio in the
emission could be easily implemented in the simulation, although the effect of
reducing the probability of contact is basically the product of the efficiency
ratios and thus such implementation is not necessary.
%
Therefore, the values of the probabilities could be modified by a constant
correction factor equal to the combined emission/reception efficiency
\citep{smith_broadcasting_2009, anchordoqui_upper_2019, forgan_collimated_2014}
or for beam-like transmissions \citep{grimaldi_signal_2017}. 
%
Other considerations include the effects of alignments, the use of stars as
sources or amplifiers \citep{Edmondson2003, borra_searching_2012}, or the
nature of the message carrier.
%
Another improvement could be the use of a spatial distribution that resembles
the spiral shape and the width of the disc of the Galaxy.
%
Regarding the distributions of the parameters, different distribution functions
for the mean lifetime of \cetis{} can be implemented as alternatives.
%
Finally, regarding the degree of idealisation, the model admits different
efficiencies for the sphere of the causal region of each \ceti{}, featuring
different searching strategies \citep{hippke_interstellar_2017}. 
%
It is also possible to consider that $D_{\rm max}$ is different for different
\cetis{}.
%
For example, a power law where powerful emissions are rare and low emissions
are common could be an improvement to the model.
%
In this work we chose not to implement this for the sake of simplicity.
%
Finally, the role of the message contents could influence on the lifespan of a
\ceti{} that receives a message, although the implementation of such behaviour
would increase the number of free parameters and would be more speculative.   
%
We limit the scope of this work to a simple version of the model. 
%
Once the model has been defined, it can be implemented as a discrete event
simulation, as described in detail in the next Section.

@@

A discrete event simulation is performed for a given model $M(\tau_a, \tau_s,
D_{\rm max})$ by keeping track of a set of variables that change each time an
event happens. 
%
In addition, the model comprises elements that are fixed for all simulations,
for example, the functional forms of the statistical distributions and the
adopted values of particular variables (see \ref{T_simu_hypotheses}).
%
The main variables that follow the evolution of the simulation are:
the positions of stars, which are sampled randomly within the GHZ; the time of
the awakening of each node (A event); and its time of disappearance (D event).
%
The variables that can be deduced from the previous ones include: the number of
\cetis{} in casual contact with at least another \ceti{} at a given time; the
number of \cetis{} as a function of time; the number of \cetis{} that receive at
least one message; the number of \cetis{} that receive a message at
least one time and successfully deliver an answer; and the number distribution
of waiting times to receive a message.
%
All these quantities are updated each time one of the four events (A, B, C, D)
occurs.

@@

Since the method is based on events, the first step is to define the
relationships between them and how events trigger changes on the state of the
system.
%
Once an event is produced, the system is updated by saving the corresponding
change in the related variables.
%
For example, a B-type event diminishes the number of active \cetis{} by one,
and also could affect the number of ongoing communications.
%
We consider a first approach to the galactic habitable zone as a 2-dimensional
annular region.
%
The adopted values for the geometry of the GHZ are 7~kpc and 9~kpc, for the
inner and outer radius, respectively \citep{lineweaver_galactic_2004}. 
%
This simple model does not take into account the variations in stellar density
given by the spiral structure of the Galaxy.
%
The main task in setting a DE simulation is to define the events and what
changes do these events produce on the system.
%
Although there are several possible approaches, we chose to follow the
evolution of the system according to the following events:

@@

We have chosen this nomenclature so that it can be easily remembered
from the initials A, B, C, D.
%
The \blackout is produced when one of the two \cetis{} halts in its
capability of emitting and receiving signals.
%
It can occur at the same time the \doomsday or before.
%
Similarly, a \ccontact can be produced at the time of the \aawakening
or later, and there can be several \contacts for the same node.
%
The time marks for each event (A and D for each \ceti{} and C and B
for each constrained causal contact) are stored as a result of the simulation.
%
Also, the list of active \cetis{} is obtained as a function of time.
%
%Some fixed variables must be set in order to carry out the simulation:
%the size of the Galactic Habitable Zone, the mean lifetime of a
%\ceti{}, $\tau_s$, the mean waiting time for the appearance of another
%\ceti{}, $\tau_a$, and the maximum distance a message can be detected.
